Brownian local times

### 2015-09-14 notes/brownian-local-times.Rmd
l

Now, let $p(x,r,\tau,A)$ denote the probability that a lineage begun at $x<0$
at a location at recombination distance $r$ from the selected site
inherits from the left-hand side of the barrier
(i.e., an ancestor on $(-\infty,0)$)
$\tau$ generations ago.
Let $p_H$ denote the density of heterozygotes in the cline
-- this should be $p_H = (1-s)/2$).
For an individual to the left of the barrier ($x<0$),
the probability of inheriting from the left of the barrier at $\tau$
is the probabiltiy that no recombination at the barrier has occurred,
plus one-half the probability that recombination in a heterozygote has occurred,
i.e., $\zeta p_H +(1-\zeta p_H)/2=(1+\zeta p_H)/2$.
If $x>0$ this is the one-half times probability a recombination in a heterozygote has occurred,
i.e., $(1-\zeta p_H)/2$.
Combining these,
the probability of ending up on the left-hand side is
$$
\begin{align*}
    p(x,r,\tau,A) = 
    \frac{1}{2} \left\{ 1 - \sgn(x) \zeta(x,r,\tau) p_H \right\} .
\end{align*}
$$

Does this actually make a cline?
```{r show_cline}
pcline <- function (x,r,tau=1,sigma=1,pH=1/2) {
    (1/2)*( 1 - sign(x)*zeta(x=x/sqrt(tau*sigma^2),r=r*tau)*pH )
}
xx <- seq(-15,15,length.out=500)
for (tau in c(5,50)) {
    plot(xx, pcline(xx,r=1), type='n', 
            xlab="distance from cline center", ylab="prob of inherited from the left", 
            main=as.expression(substitute(tau==mytau,list(mytau=tau))) )
    rr <- seq(0,0.2,length.out=11)
    pp <- sapply( rr, function (r) { pcline(xx,r=r,tau=tau) } )
    matplot(xx,pp,type='l',lty=1,col=rainbow(length(rr)),add=TRUE)
    legend("topright",lty=1,col=rainbow(length(rr)),legend=paste("r=",rr))
}
```

>>>>

Now, let $p(x,r,\tau,A)$ denote the probability that a lineage begun at $x<0$
at a location at recombination distance $r$ from the selected site
inherits from the left-hand side of the barrier
(i.e., an ancestor on $(-\infty,0)$)
$\tau$ generations ago.
(Note that $r$ should be scaled by the density and fecundity of heterozygotes in the cline --
see below.)
For an individual to the left of the barrier ($x<0$),
the probability of inheriting from the left of the barrier at $\tau$
is the probability that no recombination at the barrier has occurred,
plus one-half the probability that recombination in a heterozygote has occurred,
i.e., $\zeta +(1-\zeta)/2=(1+\zeta)/2$.
If $x>0$ this is the one-half times probability a recombination in a heterozygote has occurred,
i.e., $(1-\zeta)/2$.
Combining these,
the probability of ending up on the left-hand side is
$$
\begin{align*}
    p(x,r,\tau,A) = 
    \frac{1}{2} \left\{ 1 - \sgn(x) \zeta(x,r,\tau) \right\} .
\end{align*}
$$

Here's the cline shape:
```{r show_cline}
pcline <- function (x,r,tau=1,sigma=1) {
    (1/2)*( 1 - sign(x)*zeta(x=x/sqrt(tau*sigma^2),r=r*tau) )
}
xx <- seq(-15,15,length.out=500)
for (tau in c(5,50)) {
    plot(xx, pcline(xx,r=1), type='n', 
            xlab="distance from cline center", ylab="prob inherited from the left", 
            main=as.expression(substitute(tau==mytau,list(mytau=tau))) )
    rr <- seq(0,0.2,length.out=11)
    pp <- sapply( rr, function (r) { pcline(xx,r=r,tau=tau) } )
    matplot(xx,pp,type='l',lty=1,col=rainbow(length(rr)),add=TRUE)
    legend("topright",lty=1,col=rainbow(length(rr)),legend=paste("r=",rr))
}
```

Adjusting for cline width
=========================

Above, we were approximating the time that a reflecting lineage
spends in a small interval near zero
by the local time at zero multiplied by the width of the interval.
In fact, when the lineage is at $x$, 
the lineage recombines to the other background
in a given generation
with probability equal to the recombination rate 
multiplied by the probability of being found in a heterozygote.
If $p_{AB}(x)$ is the local fraction of heterozygotes,
and heterozygotes have relative fecundity $1-s$,
then the probability of being in a heterozygote is
$(1-s)p_{AB}(x)/((1-s)p_{AB}(x)+2p_{AA}(x))$;
if diploid genotypes are assorted randomly,
then $p_{AB} = 2p(1-p)$ and $p_AA=p^2$, so the probability that an $A$ lineage
passes through a heterozygote in a given generation is
$$
\begin{align*}
f_H(x) 
&= 
\frac{ (1-s)p(x)(1-p(x)) }{ p(x)^2 + (1-s) p(x) (1-p(x)) }
&= 
\frac{ (1-s)(1-p(x)) }{ p(x) + (1-s) (1-p(x)) } \\
&= 
\frac{ (1-s)(1-p(x)) }{ 1 - s (1-p(x)) } .
\end{align*}
$$

Bazykin's approximation to a cline is
$$
p(x) = \frac{1}{2} (1+\tanh(-2x\sqrt{s}/\sigma))
$$
Here's what that looks like:
```{r tanh_cline}
tanh_cline <- function (x,s,sigma=1) {
    (1/2)*(1+tanh(-2*x*sqrt(s)/sigma))
}
plot(xx, tanh_cline(xx,s=0.04) )
```

A lineage does something a bit more complicated,
but let's just run a random walk with drift equal to the logarithmic derivative of $p(x)$,
i.e.,
$$
\begin{align*}
    \partial_x \log(p(x)) 
    &= -\frac{2\sqrt{s}}{\sigma} 
        \frac{1-\tanh^2(-2x\sqrt{s}/σ)}{1+\tanh(-2x\sqrt{s}/σ)} .
\end{align*}
$$

```{r sim_lineage}
drift <- function (x,s,sigma=1) {
    -(2*sqrt(s)/sigma) * (1-tanh(-2*x*sqrt(s)/sigma)^2) / (1+tanh(-2*x*sqrt(s)/sigma))
}
nwalks <- 1e4
ntimes <- 1e3
s <- 0.1
# want total time to be 1, so:
eps <- 1/sqrt(ntimes)
rws <- rbind(0,matrix( rnorm(nwalks*(ntimes-1))*eps, nrow=(ntimes-1), ncol=nwalks ))
for (k in 2:ntimes) {
    rws[k,] <- rws[k-1,]+rws[k,]+drift(rws[k-1,],s)
}
layout( t(1:2), widths=c(3,1) )
matplot(rws[,1:80],type='l', ylab='position', xlab='time', ylim=range(rws) )
plot( tanh_cline(xx,s=s), xx, col='green', lwd=2, type='l', ylim=range(rws), ylab='' )
lines( (1-s)*(1-tanh_cline(xx,s=s))/(1-s*(1-tanh_cline(xx,s=s))), xx, lwd=2 )
```
